### 2017-09-26 paper/article_0626.tex
redo the modelling section a bit

In this paper we develop statistical models for the source attribution, and use a case study of human campylobacteriosis conducted in New Zealand \cite{Marsh} as a demonstration. Human campylobacteriosis is caused by \textit{Campylobacter}, of which \textit{C. jejuni} and \textit{C. coli} are the dominant species associated with 80\% and approximately 15\% of illnesses, respectively \cite{Guert}. The common symptoms of infection are diarrhea, abdominal pain and fever. However, a severe complication named Guillain Barre syndrome may develop, which is a life-threatening disease that weakens the nervous system and leads to paralysis in limbs and respiration \cite{Hanha}. The pathogen can be spread between animals, or from animals and wild birds to humans. The transmission route can be via drinking contaminated water, eating undercooked animal food products, or handling animal food products that are already contaminated by faeces. In this study, we compare the performance of the asymmetric Island model \cite{Wilso}, which considers genetic evolution when estimating the genotype sampling distribution on each source, to a simple model that utilises only the prevalence of each type to derive the sampling distribution. We then extend both models to incorporate covariates, exploring the effect of human case rurality on attribution results, modelling attribution by rurality using a linear trend or using separate categories and performing model comparison.

\section{Models for source attribution with and without information of evolutionary change}
\subsection{Notation and assumptions}
The ultimate goal of the model development is to estimate the source attribution probability, which can quantify the attribution to each source with probability given a genotype is identified in an individual. The main idea behind the establishment has two parts: the sampling distribution of genotypes associated with sources, and the proportion of human cases for each source. Specifying first the sampling distribution of genotypes found on sources is fundamental for the purpose of not only exploring how it affects the source attribution probability, but also investigating the difference in attribution effect made between the model with genetic information and the model without it.

Suppose we have isolates collected from human and non-human cases, of which $H$ isolates belong to humans, and the remaining $N$ isolates are categorized in $J$ groups as the major sources attributed to the infection. Let $I$ genotypes be the total number of unique STs detected from all isolates, and denote $n_j$ as the marginal frequency of STs found in source $j$, where $\sum_j^J n_j = N$. Typically, the number of detected STs $I$ is smaller than the sample size of isolates because some isolates may have the same genetic information.

The likelihood of observing a human case h can be found via the law of total probability, which is expressed as,

\begin{align}
  L\Big(\text{source }j; \text{ST} \Big)=\prod_{h=1}^{H}\sum_{j=1}^{J} p\Big(\text{ST}i(h) \vert \text{source } j\Big) p\Big(\text{source }j\Big),
  \label{Olikh}
\end{align}

where $p(\text{ST }i(h)\vert \text{source }j)$ is the estimated sampling distribution of genotypes for the h$^{th}$ case, $p(\text{source }j)$ is the attribution probability that a random human case is infected from source $j$. The likelihood over all human cases is multiplying each individual likelihood, which sums up the probability of detecting a genotype from the h$^{th}$ case over sources with the associated attribution probability. It is then used in the Metropolis-Hastings algorithm with a priori knowledge of $p(\text{source }j)$ to simulate samples of attribution probability. We can compare the attribution results produced from the integrated and the genetic-free models to explore the genetic efforts the integrated model contributes.

The model with genetic information used in this paper is the integrated model \cite{Wilso} adopted in the source attribution study for human campylobacteriosis \cite{Marsh}. The model has two components: an epidemiological model and an evolutionary model. The epidemiological component is a multinomial-based model capturing the number of genotypes typed from human and non-human isolates. Meanwhile, the evolutionary component estimates the rates of mutation and recombination occurring within sources, and also the migration rate at which the pathogen moves between sources by associating each source with an island. The joint model therefore makes use of the two estimates to find out how likely it is the observed genotypes detected from individuals can be attributed to sources. For simplicity, we will refer to the joint model as ``the island model'' and denote $p(\text{ST }i\vert \text{source }j, \boldsymbol{X}, \boldsymbol{G})$ as the probability the model estimates, where $\boldsymbol{X}$ represents the frequency information provided from the multinomial model and $\boldsymbol{G}$ is the parameter set of evolutionary rates evaluated in the genetic part.

To discover the genetic efforts behind the island model, a simple model is developed to produce similar inforamtion to what the island model provides. With an assumption that the sampling distribution of genotypes on each source is similar to the distribution of observed genotypes, the model is also on a multinomial basis. Let $x_{ij}$ denote the count of ST-$i$ found in source $j$ with the probability $\pi_{ij}$, where $i=1, \ldots, I$ and $j=1, \ldots, J$. To make an inference about $\pi_{ij}$, the likelihood is of a multinomial form,

\begin{align}
        L(\boldsymbol{\pi}_{j} ; \boldsymbol{x}_{j}) & = \frac{n_j!}{\prod_{i=1}^{I} x_{ij}!} \prod_{i=1}^I \pi_{ij}^{x_{ij}}, \notag
\end{align}

where $x_{ij}$ can be $0$, indicating STs are yet observed on the sources; $n_j=\sum_{i=1}^I x_{ij}$ is the total count of ST found on source $j$ and $\pi_{ij}$ is subject to $\sum_{i=1}^I \pi_{ij} =1$, and $0 \leq \pi_{ij} \leq 1$. As the family of Dirichlet distribution is a conjugate pair for the multinomial distribution, assume the prior of $\boldsymbol{\pi}_{j}$ follows a Dirichlet density with parameters $\boldsymbol{\alpha}_{j}$,

\begin{align}
  p(\boldsymbol{\pi}_{j}) & \propto \prod_{i=1}^{I}\pi_{ij}^{\alpha_{ij}-1}. \notag
\end{align}

Then, the posterior of $\boldsymbol{\pi}_{j}$ takes the form of a Dirichlet probability function with parameters $(\boldsymbol{\alpha}_{j}+\boldsymbol{x}_{j}-\boldsymbol{1})$,

\begin{align}
        p(\boldsymbol{\pi}_{j} \vert \boldsymbol{x}_{j} ) & \propto L(\boldsymbol{\pi}_{j} ; \boldsymbol{x}_{j}) p(\boldsymbol{\pi}_{j}) \notag \\ 
       & \propto \prod_{i=1}^{I}\pi_{ij}^{\alpha_{ij} + x_{ij}-1}, \quad \alpha_{ij} >0. \notag
 \end{align}

To express the belief that every human case has the same chance of getting infected from each of $J$ sources, the parameter of the Dirichlet prior is assumed as $\boldsymbol{\alpha}_{j}=\boldsymbol{1}$ in the application. Therefore, p(ST $i\vert$source $j$, $\boldsymbol{X}$) can be obtained by simulating pseudo counts for each genotype with the posterior probability $\boldsymbol{\pi}_{j}$.

Last, the overall likelihood can be computed by substituting the respective sampling distribution from the island model and the simple model for $p(\text{ST }i(h)\vert\text{source }j)$ in equation~(\ref{Olikh}).

\subsection{Campylobacteriosis data}
The source attribution study used here has MLST data collected in the Manawatu region of New Zealand from 2005 to 2014. The sampling has been done in the same time and from the same geographical location that gives 3,480 isolates taken from human and non-human cases, 1,400 of which are from humans. The non-human isolates are sampled from cattle, sheep, wild birds, environmental water and chicken carcasses, and we therefore put them into four groups of major source of infection: poultry, ruminant, water and others. The total number of unique ST typed from all isolates is 355, and 35\% of genotypes are found among human cases. Table~(\ref{tab1}) lists five common genotypes typed in human and source isolates. The first four genotypes are frequently observed in human cases. For four sources, ST-45 and ST-474 are commonly detected in poultry, while ST-42 and ST-61 are typically found in ruminants. Some studies also discover that human cases with ST-45 or ST-474 are likely attributable to poultry, whereas human cases with ST-42 or ST-61 are more likely attributable to ruminants \cite{Muell, Coll, Cart}. Furthermore, it also shows that there are no observations of ST-2381 in humans, poultry or ruminants, indicating that the genotype may only appear in the water or other sources.

>>>>

In this paper we develop statistical models for the source attribution, and use a case study of human campylobacteriosis conducted in New Zealand \cite{Marsh} as a demonstration. Human campylobacteriosis is caused by \textit{Campylobacter}, of which \textit{C. jejuni} and \textit{C. coli} are the dominant species associated with 80\% and approximately 15\% of illnesses, respectively \cite{Guert}. The common symptoms of infection are diarrhea, abdominal pain and fever. However, a severe complication named Guillain Barre syndrome may develop, which is a life-threatening disease that weakens the nervous system and leads to paralysis in limbs and respiration \cite{Hanha}. The pathogen can be spread between animals, or from animals and wild birds to humans. The transmission route can be via drinking contaminated water, eating undercooked animal food products, or handling animal food products that are already contaminated by faeces. In this study, we compare the performance of the asymmetric Island model \cite{Wilso}, which considers genetic evolution when estimating the genotype sampling distribution on each source, to a simple model that utilises only the prevalence of each type to derive the sampling distribution. We then extend both models to incorporate covariates, exploring the effect of human case rurality on attribution results, modelling attribution by rurality using a linear trend or using separate categories in a Bayesian context, and performing model comparison.

\section{Models for source attribution with and without information of evolutionary change}
\subsection{Notation and assumptions}
The ultimate goal of attribution models is to estimate the probability that the observed human cases arise from each putative source. Given genotyping information, we first estimate the sampling distribution of genotypes for each source, and then estimate the appropriate combinations of those genotype distributions that most likely give rise to that observed among human cases. Specifying first the sampling distribution of genotypes found on sources is fundamental for the purpose of not only exploring how it affects the source attribution probability, but also investigating the difference in attribution effect made between the model with genetic information and the model without it.

Suppose we have isolates collected from human and non-human cases, of which $H$ isolates belong to humans, and the remaining $N$ isolates are categorized in $J$ groups as the major sources attributed to the infection. Let $I$ genotypes be the total number of unique STs detected from all isolates, and denote $n_j$ as the marginal frequency of STs found in source $j$, where $\sum_j^J n_j = N$. Typically, the number of detected STs $I$ is smaller than the sample size of isolates as multiple isolates will be of the same type.

The likelihood of observing human cases with types $\text{ST}_1, \text{ST}_2, \ldots \text{ST}_H$ follows a multinomial distribution, which via the law of total probability may be expressed as
\begin{align}
  L\Big(\text{ST}_1, \text{ST}_2, \ldots, \text{ST}_H\Big)=\prod_{h=1}^{H}\sum_{j=1}^{J} p\Big(\text{ST}_h \vert \text{source } j\Big) p(\text{source }j),
  \label{Olikh}
\end{align}
where $p(\text{ST}_h \vert \text{source }j)$ is the probability that $\text{ST}_h$ arises from the sampling distribution of source $j$, $p(\text{source }j)$ is the attribution probability that a random human case is infected from source $j$. Given we know $p(\text{ST}_h \vert \text{source }j)$, estimation of $p(\text{source }j)$ may be found by optimising the likelihood \ref{Olikh}, for example using a Metropolis Hastings algorithm within a Bayesian context, with suitable priors on $p(\text{source }j)$.

The asymmetric Island model \cite{Wilso} adopted in the source attribution study for human campylobacteriosis \cite{Marsh} utilises the allelic profile information for each genotype in an evolutionary model, estimating mutation and recombination probabilities within, and migration probabilities between, each source `island'. It thus estimates $p(\text{ST}_h \vert \text{source }j)$ indirectly, by first estimating the evolutionary parameters, and then deriving the sampling distributions. This allows the asymmetric Island model to estimate the likelihood of observing a genotype on a source when it has not been previously observed.

To discover the effect of incorporating genetic information at the allelic profile level as used in the island model, a simple model is developed for the genotype sampling distribution. With the assumption that the observed distribution of genotypes is representative of the true distribution, we model observed genotypes using a multinomial distribution. Let $x_{ij}$ denote the count of ST-$i$ found in source $j$ with the probability $\pi_{ij}$, where $i=1, \ldots, I$ and $j=1, \ldots, J$. To make inference about $\pi_{ij}$, the likelihood is of a multinomial form,
\begin{align}
        L(\boldsymbol{\pi}_{j} ; \boldsymbol{x}_{j}) & = \frac{n_j!}{\prod_{i=1}^{I} x_{ij}!} \prod_{i=1}^I \pi_{ij}^{x_{ij}}, \notag
\end{align}
where $x_{ij}$ can be $0$, indicating STs are yet observed on the sources; $n_j=\sum_{i=1}^I x_{ij}$ is the total count of ST found on source $j$ and $\pi_{ij}$ is subject to $\sum_{i=1}^I \pi_{ij} =1$, and $0 \leq \pi_{ij} \leq 1$. As the family of Dirichlet distributions is a conjugate pair for the multinomial distribution, assume the prior for $\boldsymbol{\pi}_{j}$ follows a Dirichlet density with parameters $\boldsymbol{\alpha}_{j}$,
\begin{align}
  p(\boldsymbol{\pi}_{j}) & \propto \prod_{i=1}^{I}\pi_{ij}^{\alpha_{ij}-1}. \notag
\end{align}

Then the posterior for $\boldsymbol{\pi}_{j}$ takes the form of a Dirichlet probability function with parameters $(\boldsymbol{\alpha}_{j}+\boldsymbol{x}_{j}-\boldsymbol{1})$,
\begin{align}
        p(\boldsymbol{\pi}_{j} \vert \boldsymbol{x}_{j} ) & \propto L(\boldsymbol{\pi}_{j} ; \boldsymbol{x}_{j}) p(\boldsymbol{\pi}_{j}) \notag \\ 
       & \propto \prod_{i=1}^{I}\pi_{ij}^{\alpha_{ij} + x_{ij}-1}, \quad \alpha_{ij} >0. \notag
 \end{align}

To express the belief that every isolate is equally likely \emph{apriori}, the parameter of the Dirichlet prior is assumed as $\boldsymbol{\alpha}_{j}=\boldsymbol{1}$. Therefore, p(ST $i\vert$source $j$, $\boldsymbol{X}$) can be obtained by simulating from the Dirichlet posterior.

Lastly, the overall likelihood can be computed by substituting the respective sampling distribution from the island model and the simple model for $p(\text{ST }i(h)\vert\text{source }j)$ in equation~(\ref{Olikh}), and integrating over the posteriors.

TODO: JM: I think including an algorithm here might be useful. It could then be extended in section 3.1

\subsection{Campylobacteriosis data}

The source attribution study used here has MLST data collected in the Manawatu region of New Zealand from 2005 to 2014. The sampling has been done in the same time and from the same geographical location that gives 3,480 isolates taken from human and non-human cases, 1,400 of which are from humans. The non-human isolates are sampled from cattle, sheep, wild birds, environmental water and chicken carcasses, and we therefore put them into four groups of major source of infection: poultry, ruminant, water and others. The total number of unique ST typed from all isolates is 355, and 35\% of genotypes are found among human cases. Table~(\ref{tab1}) lists five common genotypes typed in human and source isolates. The first four genotypes are frequently observed in human cases. For four sources, ST-45 and ST-474 are commonly detected in poultry, while ST-42 and ST-61 are typically found in ruminants. Some studies also discover that human cases with ST-45 or ST-474 are likely attributable to poultry, whereas human cases with ST-42 or ST-61 are more likely attributable to ruminants \cite{Muell, Coll, Cart}. Furthermore, it also shows that there are no observations of ST-2381 in humans, poultry or ruminants, indicating that the genotype may only appear in the water or other sources.
